---
title: "fbnet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{fbnet}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center")
```

# Installation

<!-- badges: start -->
<!-- badges: end -->
At first, fbnet library should loaded. Toyped is a default example available in fbnet package. You can install fbnet like so:

``` {r, message=FALSE, results='hide'}
install.packages("fbnet") 
library(fbnet)
```

# Setting up the pedigree
"Toyped" is a pedigree example provided by fbnet package. Also, fbnet could work with pedigree object produced by pedtools R package. As could be seen, "Toyped" is a pedigree with three members (Ids 1, 2 and 3). FID is the ID, MID is the mother ID. Therefore, 1 and 2 are founders and ID 3 parents are 1 and 2.   
```{r toyped}
toyped[["pedigree"]]
```

The first step for fbnet pedigree processing is initialize the Bayesian Network through initBN function. Once generated, pbn stores some parameters, such as reference population allele frequencies. In the example two markers are analyzed, M1 and M2, with four alleles each one. 

```{r}
pbn  <- initBN(toyped)
pbn[["alelFreq"]]
```

And both the mother (ID 2) is genotyped, being M1_p, mothers paternal allele and M1_m mothers maternal allele.

```{r}
pbn[["markerEvidence"]]
```

# Building the Bayesian Network

BuildBN function perform the computations for building the Bayesian network. Query person should be defined, in this case it is the son (ID 3). Each allele of each individual represent a node in the network, edges represent conditional dependency. Using igraph package, the network could be plotted. It could be observed that the four alleles from individual 3 are conditioned by maternal alleles, paternal alleles and a linkage variable (s).

For visualizing the network igraph package should be installed and loaded.
```{r, message=FALSE, results='hide'}
install.packages("igraph")
library(igraph)
```

```{r}
bnet <- buildBN(pbn,QP=3)
plot(bnet[["DAG"]], layout=layout_with_fr, vertex.size=15,
     vertex.label.dist=2.5, vertex.label.color = "black", vertex.label.size = 24, vertex.color="black", edge.arrow.size=0.5)
```


# Computing genotype conditional probability tables

The Bayesian network allows computing allele conditional probability tables. This process could be computationally expensive. After building CPTs, variable elimination approach used, it is a process by which we successively remove variables from the Bayesian network while maintaining its ability to answer queries of interest.

```{r , message=FALSE, results='hide'}
bn1 <- buildCPTs(bnet)
resQ <- velim.bn(bn1,ordMethod="min_fill",verbose=FALSE)
```


# Analyzing conditioned probabilities

Once computed, resQ contains conditioned genotype probabilities for individual 3, based on available evidence in the pedigree. The output allows us to compare conditioned probabilities with reference population data. Firstly, some data formatting should be done.

Firstly, some packages must be loaded for data analyses and plots.
```{r, message=FALSE, results='hide'}
library(tidyverse)
library(patchwork)
library(ggplot2)
library(viridis)
library(hrbrthemes)
```

Then, resQ output could be analyzed as follows:
```{r}
Mprob1 <- resQ[["3_M1_m"]]
Fprob1 <- resQ[["3_M1_p"]]
names(Mprob1) <- c("V1", "V2")
names(Fprob1) <- c("V1", "V2")
```

Probabilities are normalized to sum 1.

```{r}
Fprob1 <- mutate(Fprob1, V2 = V2*(1/sum(Fprob1$V2)))
Fprob1 <- mutate(Fprob1, V3 = "Paternal allele")
Mprob1 <- mutate(Mprob1, V2 = V2*(1/sum(Mprob1$V2)))
Mprob1 <- mutate(Mprob1, V3 = "Maternal allele")
Pop <- as.data.frame(cbind(names(pbn[["alelFreq"]]$M1), pbn[["alelFreq"]]$M1))
Pop <- mutate(Pop, V3 = "Population")

Mprob1<-rbind(Mprob1,Pop)
Fprob1<-rbind(Fprob1,Pop)
```

Maternal allele probability for Marker 1 is strongly conditioned and differs from population. In contrast, paternally conditioned allele is not different from the population due to the lack of genetic evidence.

```{r}
p1 <- ggplot(Mprob1, 
       aes(x = V1, y = as.numeric(V2), fill= V3)) +
       geom_bar(position='dodge', stat = "identity", width=0.5) + 
       theme_classic() + xlab("Allele") + ylab("Probability")
p2 <- ggplot(Fprob1, 
       aes(x = V1, y = as.numeric(V2), fill= V3)) +
       geom_bar(position='dodge', stat = "identity", width=0.5) + 
       theme_classic() + xlab("Allele") + ylab("Probability")
p1 + p2
```


For marker 2, the same process.

```{r}
Mprob2 <- resQ[["3_M2_m"]]
Fprob2 <- resQ[["3_M2_p"]]
names(Mprob2) <- c("V1", "V2")
names(Fprob2) <- c("V1", "V2")

Fprob2 <- mutate(Fprob2, V2 = V2*(1/sum(Fprob2$V2)))
Fprob2 <- mutate(Fprob2, V3 = "Paternal allele")
Mprob2 <- mutate(Mprob2, V2 = V2*(1/sum(Mprob2$V2)))
Mprob2 <- mutate(Mprob2, V3 = "Maternal allele")
Pop2 <- as.data.frame(cbind(names(pbn[["alelFreq"]]$M2), pbn[["alelFreq"]]$M2))
Pop2 <- mutate(Pop2, V3 = "Population")

Mprob2<-rbind(Mprob2,Pop2)
Fprob2<-rbind(Fprob2,Pop2)

```

And results are plotted as follows:

```{r}
p1 <- ggplot(Mprob2, 
       aes(x = V1, y = as.numeric(V2), fill= V3)) +
       geom_bar(position='dodge', stat = "identity", width=0.5) + 
       theme_classic() + xlab("Allele Marker 1") + ylab("Probability")
p2 <- ggplot(Fprob2, 
       aes(x = V1, y = as.numeric(V2), fill= V3)) +
       geom_bar(position='dodge', stat = "identity", width=0.5) + 
       theme_classic() + xlab("Allele Marker 2") + ylab("Probability")
p1 + p2

```

# Analyzing several markers

Moreover, we can analyze and plot genotype probability matrix considering both markers. We firstly calculate the genotype probabilities for marker 1.

```{r}
M1Genopop<- as.data.frame(as.table(outer(as.numeric(Pop$V2), as.numeric(Pop$V2))))
M1Genopop<- mutate(M1Genopop,Geno = str_c(M1Genopop$Var1,"/",M1Genopop$Var2))
M2Genopop<-as.data.frame(as.table(outer(as.numeric(Pop2$V2), as.numeric(Pop2$V2)))) 
M2Genopop<- mutate(M2Genopop,Geno = str_c(M2Genopop$Var1,"/",M2Genopop$Var2))

M1_vec <- M1Genopop$Freq
names(M1_vec) <- M1Genopop$Geno
```


Same computations could be done for marker 2

```{r}
M2_vec <- M2Genopop$Freq
names(M2_vec) <- M2Genopop$Geno
GenoPop <- as.data.frame(as.table(outer(M1_vec,M2_vec)))

A <- filter(Mprob1, V3 == "Maternal allele")
B <- filter(Fprob1, V3 == "Paternal allele")
C <- filter(Mprob2, V3 == "Maternal allele")
D <- filter(Fprob2, V3 == "Paternal allele")

M1GenoPed<- as.data.frame(as.table(outer(as.numeric(A$V2), as.numeric(B$V2))))
M1GenoPed<- mutate(M1GenoPed,Geno = str_c("D","/",M1GenoPed$Var2))
M2GenoPed<-as.data.frame(as.table(outer(as.numeric(C$V2), as.numeric(D$V2)))) 
M2GenoPed<- mutate(M2GenoPed,Geno = str_c(M2GenoPed$Var1,"/",M2GenoPed$Var2))
```

And both markers genotype probabilities are computed as follows.

```{r}
M1_vec_ped <- M1GenoPed$Freq
names(M1_vec_ped) <- M1GenoPed$Geno

M2_vec_ped <- M2GenoPed$Freq
names(M2_vec_ped) <- M2GenoPed$Geno
GenoPed <- as.data.frame(as.table(outer(M1_vec_ped,M2_vec_ped)))
```

The following plots shows the probability for each genotype conditioned by the pedigree evidence or by the reference population.

```{r}
p3 <- ggplot(GenoPop, aes(x = Var2, y = Var1)) + 
  geom_raster(aes(fill=Freq)) + 
  scale_fill_gradient(low="grey90", high="blue") +
  labs(x="Marker 2", y="Marker 1", title="Population") +
  theme_bw() + theme(axis.text.x=element_text(size=5, angle=0, vjust=0.3),
                     axis.text.y=element_text(size=5),
                     plot.title=element_text(size=10))
p4 <- ggplot(GenoPed, aes(x = Var2, y = Var1)) + 
  geom_raster(aes(fill=Freq)) + 
  scale_fill_gradient(low="grey90", high="blue") +
  labs(x="Marker 2", y="Marker 1", title="Conditioned") +
  theme_bw() + theme(axis.text.x=element_text(size=8, angle=0, vjust=0.3),
                     axis.text.y=element_text(size=8),
                     plot.title=element_text(size=10))

p3 + p4
```

